---
title: "Project 7: Difference-in-Differences and Synthetic Control"
output: html_document
---


```{r}
# Install and load packages 
if (!require("pacman")) install.packages("pacman")

devtools::install_github("ebenmichael/augsynth")

pacman::p_load(# Tidyverse packages including dplyr and ggplot2 
               tidyverse,
               ggthemes,
               augsynth,
               gsynth,
               scales)

# set seed
set.seed(1)

# load data
medicaid_expansion <- read.csv('data/medicaid_expansion.csv', sep= ",")
```

# Introduction

For this project, you will explore the question of whether the Affordable Care Act increased health insurance coverage (or conversely, decreased the number of people who are uninsured). The ACA was passed in March 2010, but several of its provisions were phased in over a few years. The ACA instituted the "individual mandate" which required that all Americans must carry health insurance, or else suffer a tax penalty. There are four mechanisms for how the ACA aims to reduce the uninsured population:

- Require companies with more than 50 employees to provide health insurance.
- Build state-run healthcare markets ("exchanges") for individuals to purchase health insurance.
- Provide subsidies to middle income individuals and families who do not qualify for employer based coverage.
- Expand Medicaid to require that states grant eligibility to all citizens and legal residents earning up to 138\% of the federal poverty line. The federal government would initially pay 100\% of the costs of this expansion, and over a period of 5 years the burden would shift so the federal government would pay 90\% and the states would pay 10\%.

In 2012, the Supreme Court heard the landmark case NFIB v. Sebelius, which principally challenged the constitutionality of the law under the theory that Congress could not institute an individual mandate. The Supreme Court ultimately upheld the individual mandate under Congress's taxation power, but struck down the requirement that states must expand Medicaid as impermissible subordination of the states to the federal government. Subsequently, several states refused to expand Medicaid when the program began on January 1, 2014. This refusal created the "Medicaid coverage gap" where there are indivudals who earn too much to qualify for Medicaid under the old standards, but too little to qualify for the ACA subsidies targeted at middle-income individuals.

States that refused to expand Medicaid principally cited the cost as the primary factor. Critics pointed out however, that the decision not to expand primarily broke down along partisan lines. In the years since the initial expansion, several states have opted into the program, either because of a change in the governing party, or because voters directly approved expansion via a ballot initiative.

You will explore the question of whether Medicaid expansion reduced the uninsured population in the U.S. in the 7 years since it went into effect. To address this question, you will use difference-in-differences estimation, and synthetic control.

# Data

The dataset you will work with has been assembled from a few different sources about Medicaid. The key variables are:

- **State**: Full name of state
- **Medicaid Expansion Adoption**: Date that the state adopted the Medicaid expansion, if it did so.
- **Year**: Year of observation.
- **Uninsured rate**: State uninsured rate in that year.

# Exploratory Data Analysis

Create plots and provide 1-2 sentence analyses to answer the following questions:

- Which states had the highest uninsured rates prior to 2014? The lowest?
- Which states were home to most uninsured Americans prior to 2014? How about in 2021? **Note**: 2010 state population is provided as a variable to answer this question. In an actual study you would likely use population estimates over time, but to simplify you can assume these numbers stay about the same.

```{r}
summary(medicaid_expansion)
```

```{r}
# highest and lowest uninsured rates

high_low <-
  medicaid_expansion[order(-medicaid_expansion$uninsured_rate),] %>%
  filter(year < 2014)

high_low %>%
  ggplot(aes(x = uninsured_rate, y = State, color = State)) +
  geom_point() + 
  ggtitle('Rate of Uninsured by State prior to 2014') +
  scale_x_continuous('Uninsured Rate', labels = percent) +
  scale_y_discrete('State') +
  theme_fivethirtyeight() +
  theme(plot.title = element_text(size = 16, hjust = 0.5),
        axis.title = element_text(size = 12),
        axis.text.y = element_text(size = 5),
        legend.position = 'None')

## States with 5 highest and lowest average pre-2014 uninsured rates
high_low %>%
  group_by(State) %>%
  summarize(mean_uninsured_rate = mean(uninsured_rate)) %>%
  arrange(mean_uninsured_rate) %>%
  slice(c(1:5, 47:51)) %>%
  ggplot(aes(x = reorder(State, -mean_uninsured_rate), y = mean_uninsured_rate, fill = State)) +
  geom_col() + 
  ggtitle('Highest/Lowest Average Pre-2014 Uninsured Rates') +
  scale_y_continuous('Uninsured Rate', labels = percent) +
  scale_x_discrete('State') +
  theme_fivethirtyeight() +
  theme(plot.title = element_text(size = 16, hjust = 0.5),
        axis.title = element_text(size = 12),
        axis.text.x = element_text(hjust = 1, angle = 45),
        legend.position = 'None')

```

While the uninsured rates vary widely across time, with ranges as wide as 5 percentage points for individual states such as Wisconsin and Wyoming during the pre-2014 period, several states stand out as having particularly high and low rates of being uninsured. Nevada, Florida, New Mexico, and Texas all had uninsured rates of around 20% or higher, while the best-performing "states" of Iowa, Minnesota, DC, Hawaii, and Massachussets all had rates under 10%.

```{r}
high_low %>%
  mutate(uninsured = uninsured_rate*population) %>%
  group_by(State) %>%
  summarize(mean_uninsured = mean(uninsured)) %>%
  arrange(desc(mean_uninsured)) %>%
  slice(1:10) %>%
  ggplot(aes(x = reorder(State, -mean_uninsured), y = mean_uninsured, fill = State)) +
  geom_col() + 
  ggtitle('Largest Average Uninsured Population Pre-2014') +
  scale_y_continuous('Uninsured Population', labels = comma) +
  scale_x_discrete('State') +
  theme_fivethirtyeight() +
  theme(plot.title = element_text(size = 16, hjust = 0.5),
        axis.title = element_text(size = 12),
        axis.text.x = element_text(hjust = 1, angle = 45),
        legend.position = 'None')

## 2020 largest uninsured population
medicaid_expansion %>%
  filter(year == 2020) %>%
  mutate(uninsured = uninsured_rate*population) %>%
  arrange(desc(uninsured)) %>%
  slice(1:10) %>%
  ggplot(aes(x = reorder(State, -uninsured), y = uninsured, fill = State)) +
  geom_col() + 
  ggtitle('Largest Uninsured Populations 2020') +
  scale_y_continuous('Uninsured Population', labels = comma) +
  scale_x_discrete('State') +
  theme_fivethirtyeight() +
  theme(plot.title = element_text(size = 16, hjust = 0.5),
        axis.title = element_text(size = 12),
        axis.text.x = element_text(hjust = 1, angle = 45),
        legend.position = 'None')

```

Examining the actual populations that were uninsured during the pre-2014 time period, California, Texas, and Florida top the list due to a combination of their relatively large populations and high uninsurance rates. While Texas and Florida had among the highest average uninsured rates, California had the largest uninsured population due to its sheer population size. While the same 10 states had the largest uninsured populations in the pre-2014 period and in 2020 as a function of their large populations, California notably cut its uninsured population by more than 50%. The relatively stable size of Texas' uninsured population, on the other hand, gave it by far the largest uninsured population in 2020.

# Difference-in-Differences Estimation

## Estimate Model

Do the following:

- Choose a state that adopted the Medicaid expansion on January 1, 2014 and a state that did not. **Hint**: Do not pick Massachusetts as it passed a universal healthcare law in 2006, and also avoid picking a state that adopted the Medicaid expansion between 2014 and 2015.
- Assess the parallel trends assumption for your choices using a plot. If you are not satisfied that the assumption has been met, pick another state and try again (but detail the states you tried).

```{r}
# Parallel Trends plot

#California as state adopted 1/1/2014
#Utah as state that didnt adopt until 1/1/2020

medicaid_expansion %>%
  filter(State %in% c("California", "Utah")) %>%
  ggplot() + 
  geom_point(aes(x = year, 
                 y = uninsured_rate, 
                 color = State)) +
  geom_line(aes(x = year, 
                y = uninsured_rate, 
                color = State)) +
  geom_vline(aes(xintercept = 2014)) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  ggtitle('California and Utah Uninsured Rate \n  before/after 2014 Medicaid expansion') +
  xlab('Year') +
  ylab('State Uninsured')

# the two states do not move in parallel pre-2014; will pick Nebraska (didn't adopt until 1/10/2020) and see what happens

medicaid_expansion %>%
  filter(State %in% c("California", "Utah", "Nebraska")) %>%
  ggplot() + 
  geom_point(aes(x = year, 
                 y = uninsured_rate, 
                 color = State)) +
  geom_line(aes(x = year, 
                y = uninsured_rate, 
                color = State)) +
  geom_vline(aes(xintercept = 2014)) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  ggtitle('California, Utah, and Nebraska Uninsured Rate \n  before/after 2014 Medicaid expansion') +
  xlab('Year') +
  ylab('State Uninsured')

```


- Estimates a difference-in-differences estimate of the effect of the Medicaid expansion on the uninsured share of the population. You may follow the lab example where we estimate the differences in one pre-treatment and one post-treatment period, or take an average of the pre-treatment and post-treatment outcomes

```{r}
# Difference-in-Differences estimation


# pre-treatment difference
pre_diff <- medicaid_expansion %>%
  filter(State %in% c("California","Nebraska")) %>%
  filter(year == 2013) %>%
  select(State, 
         uninsured_rate) %>% 
  spread(State, 
         uninsured_rate) %>%
  summarise(California - Nebraska)
  
# post-treatment difference
post_diff <- medicaid_expansion %>%
  filter(State %in% c("California","Nebraska")) %>%
  filter(year == 2015) %>%
  select(State, 
         uninsured_rate) %>%
  spread(State, 
         uninsured_rate) %>%
  summarise(California - Nebraska)

# diff-in-diffs
diff_in_diffs <- post_diff - pre_diff
diff_in_diffs
```


## Discussion Questions

- Card/Krueger's original piece utilized the fact that towns on either side of the Delaware river are likely to be quite similar to one another in terms of demographics, economics, etc. Why is that intuition harder to replicate with this data?
- **Answer**: There is likely a lot of heterogeneity between the states in terms of demographics, SES, other covariates.  This would violate our parallel trends assumption.  Essentially we want to make the change in the uninsured rate  in the control group (here Nebraska) to be a good proxy for the counterfactual change in uninsured rate in our treatment group (here California) had they not been treated (i.e expanded Medicaid). This is less likely to be the case if there are a variety of factors that make the two states different.

- What are the strengths and weaknesses of using the parallel trends assumption in difference-in-differences estimates?
- **Answer**: 
Pros: 
- Easily interpretable 
- The assumption allows you to use just one pre- and one post-treatment period
- You are able to use smaller time periods where difference between treatment and control groups are constant 

Cons:
- There is no statistical test for this assumption
- Is less likely to hold with longer time periods
- Violation of this assumption will lead to biased estimates

# Synthetic Control

Estimate Synthetic Control

Although several states did not expand Medicaid on January 1, 2014, many did later on. In some cases, a Democratic governor was elected and pushed for a state budget that included the Medicaid expansion, whereas in others voters approved expansion via a ballot initiative. The 2018 election was a watershed moment where several Republican-leaning states elected Democratic governors and approved Medicaid expansion. In cases with a ballot initiative, the state legislature and governor still must implement the results via legislation. For instance, Idaho voters approved a Medicaid expansion in the 2018 election, but it was not implemented in the state budget until late 2019, with enrollment beginning in 2020.

Do the following:

- Choose a state that adopted the Medicaid expansion after January 1, 2014. Construct a non-augmented synthetic control and plot the results (both pre-treatment fit and post-treatment differences). Also report the average ATT and L2 imbalance.

```{r}
# non-augmented synthetic control
```

- Re-run the same analysis but this time use an augmentation (default choices are Ridge, Matrix Completion, and GSynth). Create the same plot and report the average ATT and L2 imbalance.

```{r}
# augmented synthetic control
```

- Plot barplots to visualize the weights of the donors.

```{r}
# barplots of weights
```

**HINT**: Is there any preprocessing you need to do before you allow the program to automatically find weights for donor states?

## Discussion Questions


- What are the advantages and disadvantages of synthetic control compared to difference-in-differences estimators?
- **Answer**: 

- One of the benefits of synthetic control is that the weights are bounded between [0,1] and the weights must sum to 1. Augmentation might relax this assumption by allowing for negative weights. Does this create an interpretation problem, and how should we balance this consideration against the improvements augmentation offers in terms of imbalance in the pre-treatment period?
- **Answer**:

# Staggered Adoption Synthetic Control

## Estimate Multisynth

Do the following:

- Estimate a multisynth model that treats each state individually. Choose a fraction of states that you can fit on a plot and examine their treatment effects.

```{r}
# multisynth model states
```

- Estimate a multisynth model using time cohorts. For the purpose of this exercise, you can simplify the treatment time so that states that adopted Medicaid expansion within the same year (i.e. all states that adopted epxansion in 2016) count for the same cohort. Plot the treatment effects for these time cohorts.

```{r}
# multisynth model time cohorts
```

## Discussion Questions

- One feature of Medicaid is that it is jointly administered by the federal government and the states, and states have some flexibility in how they implement Medicaid. For example, during the Trump administration, several states applied for waivers where they could add work requirements to the eligibility standards (i.e. an individual needed to work for 80 hours/month to qualify for Medicaid). Given these differences, do you see evidence for the idea that different states had different treatment effect sizes?
- **Answer**: 

- Do you see evidence for the idea that early adopters of Medicaid expansion enjoyed a larger decrease in the uninsured population?
- **Answer**: 

# General Discussion Questions

- Why are DiD and  synthetic control estimates well suited to studies of aggregated units like cities, states, countries, etc?
- **Answer**:

- What role does selection into treatment play in DiD/synthetic control versus regression discontinuity? When would we want to use either method?
- **Answer**: